### 2016-02-11 reviews/answer_to_revieweres.txt

> Line 173-181: I wonder whether this explanation is appropriate for this audience.

ANSWER: we made it a bit shorter, but not too much so; we think it is helpful for the reader to be reminded at least of the basic steps of a method

> Line 175: Were the eigenvectors of M’M and MM’ merged to arrive at the ‘first k eigenvectors’?

ANSWER: the columns of matrix U contain the eigenvectors of MM' (n x n matrix), the columns of matrix V contain the eigenvectors of M'M (m x m matrix). The n diagonal elements of matrix Sigma (n x m) are the square root of the eigenvalues of matrix U (n x n) (which should correspond also to the non-zero eigenvalues of V' / M'M). The first k eigenvectors of U were then determined based on the ordered corresponding eigenvalues. The matrix U* (n x k) contains these k eigenvectors, which multiplied by the k regression coefficients in vector beta return the (n x 1) column vector of estimated genotypes at locus j.

### 2016-02-11 reviews/answer_to_revieweres.txt

> Line 440: The reasoning on factor ii) is not clear – would the results for rice be less accurate when analysing the genome at once? Also, why would it simplify the imputation process?

ANSWER: First, the per-chromosome analysis simplifies calculations, and allow for straightforward parallelization. Additionally, the results with rice as a whole are less accurate when markers are shuffled [add numbers/supplementary Table/Figure?]; this is because the LD structure is better preserved when reshuffling markers.

> Line 443: Did you explore the population structure of the three datasets? Would presence/absence of (severe) population structure affect imputation accuracy?

Answer: There was no population structure in alfalfa (see Annicchiarico et al. 2015). On the other hand, there was population structure in rice (indica, tropical and temperate japonica). Most imputation methods implicitly exploit population structure (e.g. KNN --> distance based on genetic relatedness; Beagle --> haplotype reconstructed based on genetic similarities), without explicitly modelling it.  Imputation of missing genotypes is not GWAS, where the significance and estimate of SNP effects may be inflated if population structure is not included in the model), and the impact on the accuracy of imputation is likely to be small. In cattle from Scandinavian countries, Brondum et al (2012) actually found higher imputation accuracy when combining populations (cattle breeds) in Beagle (without explicitly modelling the population structure), most likely as a result of the larger sample size. However, the influence of population structure on the accuracy of imputation was not formally tested in this work. Though interesting, we feel that this is outside of the scope of the present paper, which already contains a large amount of comparisons and results. These considerations were added to the discussion.

### 2016-02-11 reviews/answer_to_revieweres.txt

General Comments:
> The four methods used for imputation are now very common. Even software tool like Tassel host imputations methods like FILLIN, FSFHap, Numerical, and Beagle. There are several methods and tools available for imputation and many studies demonstrating comparisons have been published.

ANSWER: the main interest in our work was to assess the imputation efficiency in two important plant species, rice (major crop species) and alfalfa (forage species). We wanted to evaluate the effects of increasing missing rates, of aligning markers along the genome, and of the size of the dataset. We were also interested in checking how different imputation methods responded to different conditions. The main points that we think are novel and of interest in our owrk are: i) the imputation of missing data from GBS experiments in alfalfa (M. sativa) and rice (O. sativa); ii) the breakdown of the imputation accuracy into the different genotype classes (major and minor homozygotes, heterozygotes), never done in other comparable papers; iii) the effect of aligning markers to the reference genome of the species or to that of a closely related species; iv) the effect of shuffling marker position, that allows to quantify the importance of having (good) alignments.

> More particularly, in species without availability of whole genome sequence methods for the imputation of SNP data have been published. For instance, Ward, et al. (Saturated linkage map construction in Rubus idaeus using genotyping by sequencing and genome-independent imputation. 2013. BMC genomics, 14(1), 2).
> Even in Alfaalfa (Medicago sativa) Rocher et al. (Validation of Genotyping-By-Sequencing Analysis in Populations of Tetraploid Alfalfa by 454 Sequencing. 2015. PloS one, 10(6), e0131918.) demonstrated imputation at missing rate 0.5.

ANSWER: the mentioned papers are indeed relevant, and were cited in the introduction of the paper. However, none of the cited papers focus on algorithm comparison, nor examine the effect of different levels of call rate and missing rate.

@@

ANSWER: the title was changed to "Marker imputation efficiency for Genotyping-By-Sequencing data in rice (Oryza sativa) and alfalfa (Medicago sativa)"

> Abstract need to be more concise and informative
ANSWER: we rephrased parts of the abstract.

Introduction:
-------------
> Line 38-39, GWAS can not handle missing data- need citation and reframe the sentence. Better to verify the fact.

ANSWER: We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics, and cited other common GWAS and GS software libraries (BGLR, rrBLUP, GenABEL) that refuse input containing missing data points or that contain imputation functionalities to prepare data for analysis. Especially linear algebra and arithmetic methods suffer from missing data, whereas numerical methods like MCMC can better cope with them. In any case, using missing data, is not impossible: e.g. maximum likelihood (Handling missing data by maximum likelihood, by P.D. Allison, 2012; or Schafer, 1997: Analysis of incomplete multivariate data). We feel this is out of the scope of the present paper, though, and did not indulge in such speculations. We simply modified the sentence and added a reference.
We rephrased the sentence to highlight that imputation is often a necessary step.

@@

Manuscript: Marker imputation efficiency for Genotyping-By- Sequencing data of crop species with and without a reference genome
Authors: Nelson Nazzicari, Filippo Biscarini, Paolo Cozzi, E. Charles Brummer, Paolo Annicchiarico

General comments
This manuscript deals with an important and emerging topic in plant (and animal) breeding as GBS is considered as a viable alternative to SNP array-based genotyping. These GBS data come with major challenges, also on the imputation of missing values. This manuscript describes the application of six algorithmic approaches to data on two plant crops that are on the edges of the wide range of genotype complexity in plants. Not surprisingly, the Beagle software performs very well (superior over all other methods) when map information is available, and much worse, when map information is absent. 
My main concerns on the paper are:

> 1. Why these two species? They have very different properties which makes it difficult to associate the differences in results to particular characteristics. For example, is ploidy level more critical than a reference genome?  It might have been more insightful when including an outbred diploid species with a reference genome and similar to the rice to assess imputation accuracy with and without reference genome information.

ANSWER: we decided to use two very different species to cover a wide range of possible outcomes. The core idea is to follow the well known practice of having an easy benchmark and a difficult one, as it is often done in machine learning community. Moreover, while a general study with several species and datasets would be desirable, it was out of the scope of the current work (and effort). The paper results are consistent both internally and with the cited literature, and we think shine some light on the examined topic, even without the use of many different datasets.
We however understand the reasoning in the reviewer's question, and decided to adjust the declared target of the paper to be less general and focus more on the studied species. The change of title goes in that direction, as much as a big part of the updated text.

@@

Specific comments
> Line 52: It would be nice if you could provide a list of associated packages, AND explain why Beagle was selected out of these methods.

ANSWER: we added several references to software sources.

> Line 78: There is no development of the best method in this study, nevertheless the results of this study may identify the essential properties of such a method. Was this discussed?

ANSWER: given the very different mathematical models underlying the examined algorithms (or all imputation algorithms, for what it's worth) we think that the essential properties of a good imputation algorithm are already discussed, namely: accuracy (total and per class); resilience to growing missing level; computation times.
All these aspects have been thoroughly discussed.

@@

> Line 134: Why were these values (4 and 11) were chosen? I do not see the rational of taking 4 for a heterozygote and 11 for homozygote - shouldn’t these be the reverse?

ANSWER: this is a common practice in GBS SNP calling. We updated the text, added references to other works doing the same, and added a section in supplementary material explaining a bit of the related math. The short answer is that no, it's in general more difficult to call a homozygote, so you want to have more reads.

> Line 135-139: The reduction from 5 genotypic classes to 3 genotypic classes in alfalfa is a critical simplification step. There is no justification why this step was executed. My assumption is that this is because of the (software) methods that were used. The explanation and justification should be provided here. I am wondering whether the four general methods could be applied to the 5 genotypic classes – was this considered/studied?

ANSWER: the 5-to-3 reduction is not strictly related to algorithms. In fact, the general algorithms would work easily on any number of classes: MNI, SVDI and RFI would produce continuous values in the [0-4] range, and KNNI is naturally oriented toward a discrete outputs. Beagle and FILLIN, on the other hand, are less flexible, and using them on tetraploid genotypes is not straightforward.
However, the main reason for the simplification of heterozygotes is the available GBS read depth. To be able to call the heterozygote dosage a high number of reads per locus is required, and accepting only the markers with enough reads highly reduces their number. We explained the choice in the proper M&M section and added reference to papers following this (common) practice.

@@

> Line 401-407: This is still rather descriptive without any postulation whether any of the methods are less or more sensitive to MAF and/or balance.  The superior results of Beagle in rice are mostly driven by utilizing map/assembly information which is ignored by other methods.

ANSWER We updated the section, improving the description and commenting the results of an extra method that uses marker information.

> Line 440: The reasoning on factor ii) is not clear – would the results for rice be less accurate when analysing the genome at once? Also, why would it simplify the imputation process?

ANSWER: First, the per-chromosome analysis simplifies calculations, and allows for straightforward parallelization. Additionally, the results with rice as a whole are less accurate when markers are shuffled. In fact, accuracy losses are:

- total: -6% on total rice, -3% on chromosomes
- major homozigote: -2% on total rice, -1% on chromosomes
- minor homozigote: -41% on total rice, -22% on chromosomes

This happens because the LD structure is better preserved when reshuffling markers within chromosomes.

@@

> Line 443: Did you explore the population structure of the three datasets? Would presence/absence of (severe) population structure affect imputation accuracy?

ANSWER: we updated the manuscript to discuss the issue of population structure. In brief:

i) there is no population structure in alfalfa, as reported by published study. Rice has structure, though, since the dataset is a collation of several subpopulations (indica/tropical/temperate)
ii) population structure is implicitly exploited by most imputation methods, in particular those performing data clusterization (e.g. KNN --> distance based on genetic relatedness) 
iii) it is however not likely to affect much imputation accuracy. This is not GWAS, and the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed.

@@

> Line 450: It is too easy to state that there is additional complexity to make use of a reference genome of a related species. It would be of HIGH relevance to explore this as it may be applied in many species that lack a reference genome themselves.

ANSWER: we added the study using Medicago Truncatula.

> Line 535-540: Too general/vague and thus not clear why this paragraph is included in the conclusion section (or even in the manuscript)!

ANSWER: We updated the section to provide a more incisive conclusion.

> Figure 1: Be consistent in using rates versus percentages, i.e., this Figure uses rates while on Line 311 you refer to percentages.

### 2016-02-11 inputs/abstract.tex
several adds after afternoon pass

\begin{abstract}
Genotyping-by-sequencing (GBS) is a rapid and cost-effective genome-wide 
genotyping technique applicable whether a reference genome is available 
or not. Due to the cost-coverage trade-off, however, GBS typically produces 
large amounts of missing marker genotypes, whose imputation becomes therefore 
both challenging and critical for later analyses.\\
In this work, the performance of four general imputation methods (K-Nearest 
Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
and two genotype-specific method (``Beagle'' and FILLIN) were measured 
on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 
without reference genome) and 
rice (\emph{Oryza sativa}, diploid, 100\% homozygous, with reference genome). Benchmarks 
consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
proportions (up to 20\%) of known genotypes masked for imputation. The 
relative performance was measured as the total proportion of correctly imputed 
genotypes, globally and within each genotype class (two homozygotes in rice, 
two homozygotes and one heterozygote in alfalfa). Imputation efficiency was 
measured with computation times.\\
In general imputation accuracy was robust to increasing 
missing rates, and consistently higher in rice than in alfalfa. Accuracy 
was as high as 90-100\% for the major (most frequent) homozygous genotype, but dropped 
to 80-90\% (rice) and below 30\% (alfalfa) on the minor homozygous genotype. 
Beagle was the best performing method, both accuracy- 
and time-wise, when using a reference genome. When no reference 
genome was available, KNNI and RFI gave the highest accuracies, with KNNI 
being much faster.

>>>>

\begin{abstract}
Genotyping-by-sequencing (GBS) is a rapid and cost-effective genome-wide 
genotyping technique applicable whether a reference genome is available 
or not. Due to the cost-coverage trade-off, however, GBS typically produces 
large amounts of missing marker genotypes, whose imputation becomes therefore 
both challenging and critical for later analyses.\\
In this work, the performance of four general imputation methods (K-Nearest 
Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
and two genotype-specific methods (``Beagle'' and FILLIN) were measured 
on GBS data from alfalfa (\emph{Medicago sativa L.}, autotetraploid, heterozygous, 
without reference genome) and 
rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Benchmarks 
consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
proportions (up to 20\%) of known genotypes masked for imputation. The 
relative performance was measured as the total proportion of correctly imputed 
genotypes, globally and within each genotype class (two homozygotes in rice, 
two homozygotes and one heterozygote in alfalfa).\\
we found that imputation accuracy was robust to increasing 
missing rates, and consistently higher in rice than in alfalfa. Accuracy 
was as high as 90-100\% for the major (most frequent) homozygous genotype, but dropped 
to 80-90\% (rice) and below 30\% (alfalfa) in the minor homozygous genotype. 
Beagle was the best performing method, both accuracy- 
and time-wise, when using a reference genome. When no reference 
genome was available, KNNI and RFI gave the highest accuracies, with KNNI 
being much faster.

### 2016-02-11 inputs/conclusions.tex
several adds after afternoon pass

\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome. In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency. The alignment of markers to closely related species can help, but comes at the price of discarding the non-aligning markers. Moreover, performances indicate that general methods are still the best option.\\
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\
Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.

>>>>

\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome. In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency. The alignment of markers to closely related species can help, but comes at the price of discarding markers that do not align, and with general methods still performing better.\\
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\

\textcolor{red}{TO BE DELETED?\\Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.}

Exploring ways to improve imputation accuracy in GBS data is an important research area and can help making genotyping-by-sequencing a very cost-effective technique. In this paper we shown how general imputation algorithms are a valid alternative to genotype-oriented ones when reference genomes are not available or too expensive in terms of unaligned markers. Moreover, we highlighted the importance of examining the accuracy on minority classes, since the common practice of summarizing performances in a single index can easily mislead to undesired results.

### 2016-02-11 inputs/discussion.tex
several adds after afternoon pass

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. On our data sets we found missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation error can have a negative impact on successive analyses (e.g. on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.

@@

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling phase. This semplification step brought alfalfa dataset farther from biology;\\
iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process;\\
iv) rice markers were aligned on native genome, while alfalfa markers were aligned on a different species.

Alfalfa datasets did not present population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulations. Population structure is implicitly used by algorithms like Beagle, FILLIN and KNNI, that create clusters of similar genotypes, and it is likely to affect imputation accuracy.\\
The most dramatic difference between datasets was found with FILLIN. The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms). This behaviour is probably due to FILLIN being designed to target inbred lines (as the name itself indicates).

\subsection{Reference genome assembly and ordered vs unordered markers}
\label{sec:reference_genome_ordered_vs_unordered}
When a reference genome is available, markers positions can be used in the imputation process. All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.

In rice, where a reference genome is available, we could assess the effect of using ordered vs. unordered markers for imputation. For alfalfa there is no reference genome sequenced yet. The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle used. However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align. In our case only 57.54\% (Alfalfa-Med) and 57.86\% (Alfalfa-PV) aligned on the \emph{M. Truncatula} genome, compared to the 88.66\% of rice markers aligning on \emph{O. Sativa} genome, a loss of markers that could or could not be acceptable, depending on the specific application requirements. \\
Among the imputation methods used in this study, Beagle and FILLIN are the only ones that make use of marker information, and were therefore tested with ordered and randomly shuffled markers. In rice dataset with ordered markers Beagle showed astounding resilience to high missingness, with imputation accuracy in the minority class $>$95\% even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 75\% (10\% allowed, 1\% artificial missing rates) to almost 50\% (70\% allowed, 20\% artificial missing rates). When tested with alfalfa Beagle largerly overestimated by the majority class and had very poor performances on heterozygous class. Shuffling marker positions further worsened the accuracy, and almost all missing data point assigned to the majority class.\\
FILLIN performances on rice follow the same pattern and worsen substantially when markers are shuffled. This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.

>>>>

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. On our data sets we found missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation errors can have a negative impact on successive analyses (e.g. on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.

@@

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling. This semplification step made alfalfa data less adherent to the underlying biology;\\
iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa). This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on a different species.

Alfalfa datasets did not present population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulations. Population structure is implicitly used by algorithms like Beagle, FILLIN and KNNI, that create clusters of similar genotypes, and it is likely to affect imputation accuracy.\\
The most dramatic difference between datasets was found with FILLIN. The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms). This behaviour is probably due to FILLIN being specifically designed to target inbred lines (as the name itself indicates).

\subsection{Reference genome assembly and ordered vs unordered markers}
\label{sec:reference_genome_ordered_vs_unordered}
When a reference genome is available, marker position can be used in the imputation process. All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.

In rice, where a reference genome is available, we could assess the effect of using ordered vs. unordered markers for imputation. For alfalfa there is no reference genome sequenced yet. The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle be used. However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align. In our case only 57.54\% (Alfalfa-Med) and 57.86\% (Alfalfa-PV) aligned on the \emph{M. Truncatula} genome, compared to the 88.66\% of rice markers aligning on \emph{O. Sativa} genome. This left with 23\,438 and 18\,923 SNPs to be used for analysis with ordered markers. \\
Among the imputation methods used in this study, Beagle and FILLIN are the only ones that make use of marker information, and were therefore tested with ordered and randomly shuffled markers. In rice dataset with ordered markers Beagle showed astounding resilience to high missingness, with imputation accuracy in the minority class $>$95\% even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 75\% (10\% allowed, 1\% artificial missing rates) to almost 50\% (70\% allowed, 20\% artificial missing rates). When tested with alfalfa Beagle largerly overestimated the majority class and had very poor performances on the heterozygous class. Shuffling marker positions further worsened Beagle performance in alfalfa, and almost all missing genotypes were assigned to the majority class.\\
FILLIN performance in rice followed the same pattern and was substantially worse when markers were shuffled. This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.

### 2016-02-11 inputs/introduction.tex
several adds after afternoon pass

been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015})
and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}). These studies

@@

alternative imputation models.\\

@@

implemented in the software package Beagle, and a method of haplotype
reconstruction around recombination sites, implemented in the FILLIN algorithm
as part of Tassel suite). The four general algorithms were chosen as well known 

>>>>

been reported in alfalfa (\emph{Medicago sativa L.}, \cite{Rocher_validation_2015})
and red raspberry (\emph{Rubus idaeus L.}, \cite{ward_saturated_2013}). These studies

@@

alternative imputation methods.\\

@@

implemented in the software package Beagle, and the method of based on haplotype
reconstruction around recombination sites implemented in the FILLIN algorithm
as part of the Tassel suite). The four general algorithms were chosen as well known 
